---
title: "Lab 1 report"
author: "Simon Jorstedt & Siddhesh Sreedar"
date: "`r Sys.Date()`"
output: pdf_document
---


# Question 1

```{r}
# Problem setup

## Data parameter values
theta <- seq(-4, 4, 0.01)
x <- c(-2.8, 3.4, 1.2, -.3, -2.6)

# Log-likelihood function
loglike <- function(theta){-5*log(pi)-sum(log(1+(x-theta)^2))}

# Log-likelihood first derivative function
loglike_derivative <- function(theta){sum(2*(x-theta)/(1+(x-theta)^2))}
```

## Subquestion a
> Plot the log likelihood function for the given data in the range from -4 to 4. Plot the derivative in the same range and check visually how often the derivative is equal to 0.

```{r}
# Log likelihood plot
y_a1 <- c()
for (theta_i in theta){
  y_a1 <- c(y_a1, loglike(theta_i))
}

plot(theta,
     y_a1,
     type="l",
     main = "Log likelihood",
     xlab = "theta",
     ylab = "value")

# Log likelihood derivative plot
y_a2 <- c()
for (theta_i in theta){
  y_a2 <- c(y_a2, loglike_derivative(theta_i))
}

plot(theta,
     y_a2,
     type="l",
     main = "Log likelihood derivative",
     xlab = "theta",
     ylab = "value")
abline(b=0, a=0)
```

The derivative crosses y=0 three times.

## Subquestion b

```{r}
# Bisection function

bisect_estimate <- function(a, b, epsilon = 0.000001){
  # a is the left start guess
  # b is the right start guess
  # epsilon is the desired convergence error
  
  y1 <- loglike_derivative(a)
  y2<- loglike_derivative(b)
  stopifnot("Invalid boundaries `a` and `b`" = y1 * y2 < 0)
  
  x_t <- c()
  # In the beginning, x_t is either empty or only has one value.
  # When x_t has at least two values, evaluate the last two values with
  # respect to the convergence error.
  while(length(x_t) < 2 || abs(tail(x_t,1)-tail(x_t,2)[1]) > epsilon){
    y1 <- loglike_derivative(a) #sum(2*(x-a)/(1+(x-a)^2))
    y2<- loglike_derivative(b) #sum(2*(x-b)/(1+(x-b)^2))
    
    # Calculate new boundary
    x_t <- c(x_t,(a + b)/2)
    
    if(loglike_derivative(a)*loglike_derivative(tail(x_t,1)) <= 0){
      b <- tail(x_t,1)
    }
    if(loglike_derivative(b)*loglike_derivative(tail(x_t,1)) < 0){
      a <- tail(x_t,1)
    }
  }
  cat("Estimate of x_value: ", tail(x_t, 1))
}
```

## Subquestion c

```{r}

```

The global maximum is achieved at about x = -0.2952452

## Subquestion d

```{r}

values<-seq(-4,4,0.1)

first_der_values<-data.frame("value"= NA,"derivative" =NA)

for (i in 1:length(values)){
  first_der_values[i,1]<-values[i]
  first_der_values[i,2]<-loglike_derivative(values[i])
  
}

pos<-subset(first_der_values,derivative>0)
neg<-subset(first_der_values,derivative<0)

final<-c()

len<-nrow(pos)
len2<-nrow(neg)

for (i in 1:len){
  for( j in 1:len2){
    
    if(pos[i,1]<neg[j,1]){
      a<-pos[i,1]
      b<-neg[j,1]
    }else{
      b<-pos[i,1]
      a<-neg[j,1]
    }
    x_i <- c()
    
    while(length(x_i) < 2 || abs(tail(x_i,1)-tail(x_i,2)[1])>0.00000001){
      y1 <- loglike_derivative(a) #sum(2*(x-a)/(1+(x-a)^2))
      y2<- loglike_derivative(b) #sum(2*(x-b)/(1+(x-b)^2))
      
      if(length(x_i) == 0 && y1 * y2 >= 0){print("Break");break}
      
      x_i <- c(x_i,(a + b)/2)
      
      if(loglike_derivative(a)*loglike_derivative(tail(x_i,1)) <= 0){
        b <- tail(x_i,1)
      }
      if(loglike_derivative(b)*loglike_derivative(tail(x_i,1)) < 0){
        a <- tail(x_i,1)
      }
    }
    final<-c(final,(tail(x_i, 1)+tail(x_i, 2)[1])/2)
  }
}

plot(theta, y_a1, type="l")
abline(v=final)


plot(theta, y_a2, type="l")
abline(b=0, a=0)
abline(v=final)


final2<-as.data.frame(final)


#-1.6017752,-0.2952455,-2.0821239

y_a11 <- c()
optim_val<-c(-1.6017752,-0.2952455,-2.0821239)
for (theta_i in optim_val){
  y_a11 <- c(y_a11, -5*log(pi)-sum(log(1+(x-theta_i)**2)))
}
#[1] -13.73572 -13.40942 -13.70770

#So the highest value is when theta is -0.2952455

```


# Question 2

## Subquestion a

```{r}
# Variance estimation function
myvar <- function(x){
  output <- (sum(x^2) - sum(x)^2 / length(x)) / (length(x)-1)
  
  #cat("Variance estimate:", output, "\n")
  return(output)
}
```

## Subquestion b

```{r}
x <- rnorm(n=10^4, mean=10^8, sd=1)
```

## Subquestion c

```{r}
# The difference for each subset


myvar_vector <- c()
var_vector <- c()
for (i in 1:10^4){
  myvar_vector <- c(myvar_vector, myvar(x[1:i]))
  var_vector <- c(var_vector, var(x[1:i]))
  #difference <- c(difference, myvar(x[1:i])-var(x[1:i]))
}
difference <- myvar_vector - var_vector

plot(difference, type="p")
plot(myvar_vector, var_vector)
```


## Subquestion d

```{r}
myvar_improved_1<-function(x){
   m <- mean(x)
  sum((x-m)**2)/(length(x)-1)
}

myvar_improved_2<-function(x){
  var(x)*(length(x)-1)/length(x)
  
}

#checking difference with the first improvement
myvar_improved_1_vector <- c()
var_vector <- c()
for (i in 1:10^4){
  myvar_improved_1_vector <- c(myvar_improved_1_vector, myvar_improved_1(x[1:i]))
  var_vector <- c(var_vector, var(x[1:i]))
}

difference1 <- myvar_improved_1_vector - var_vector


#checking difference with the seconf improvement
myvar_improved_2_vector <- c()
var_vector <- c()
for (i in 1:10^4){
  myvar_improved_2_vector <- c(myvar_improved_2_vector, myvar_improved_2(x[1:i]))
  var_vector <- c(var_vector, var(x[1:i]))
}

difference2 <- myvar_improved_2_vector - var_vector



#myvar_improved_1 gives better results and close to the var() function in R. 

```
