---
title: "Q1"
author: "Simon Jorstedt"
date: "`r Sys.Date()`"
output: pdf_document
---


```{r}
# Setup packages and data

library(magrittr)
library(dplyr)

# Read lottery data
df_lottery <- read.table("lottery.txt", col.names = c("Draft_No", "Day_of_year", "Letter_rank"))

```

```{r, echo=F}
# Question 1.1

plot(y=df_lottery$Draft_No,
     x=df_lottery$Day_of_year,
     main = "Fig 1.1: Draft Num. vs Day of year")

```

In figure 1.1, there is no immediate pattern that emerges. The spread of draft numbers appears roughly uniform.

```{r, echo=F}
# Question 1.2

# Fit a regular linear model
linmod_curve <- lm(data = df_lottery,
                   Draft_No ~ Day_of_year)

# Fit a loess model
loess_curve <- loess(data = df_lottery,
                     Draft_No ~ Day_of_year)

# Add loess fitted values to dataframe
df_lottery <- df_lottery %>% mutate(loess_fitted = loess_curve$fitted)

# Plot data along with both fitted curves
p_linmod_curve <- plot(y=df_lottery$Draft_No,
                       x=df_lottery$Day_of_year,
                       main = "Fig 1.2: Draft Num. vs Day of year with")
# Add lm line
abline(b = linmod_curve$coefficients[2], a = linmod_curve$coefficients[1], col="red", lwd=2)

# Add loess curve
loess_data <- df_lottery %>% arrange(Day_of_year)
points(x=loess_data$Day_of_year, loess_data$loess_fitted, type = "l", col="blue", lwd = 2)

```

In Figure 1.2, we see that the normal linear regression line and the loess curve behave very similarly. They are both almost horizontal, which imples a low correlation between the variables. The slight negative slopes implies that higher draft numbers are ever so slightly more likely for the early days of the years. The lottery is surely random, but is it truly uniform?





