---
title: "Untitled"
output: html_document
date: "2023-12-19"
---

```{r include=FALSE}
library(compositions)
```

Question-1

I have structured the code in such a way that we proceed with doing all the part for each encoding before going to the next encoding

Encoding-1 
```{r echo=TRUE}

### 1)

### a)
encoding1<-function(n){
  data<-list()
  for(i in 1:n){
    data[[i]]<-c(sample(n,1),sample(n,1))
    while(any(duplicated(data))== TRUE){
      data[[which(duplicated(data))]]<-c(sample(n,1),sample(n,1))
    }
  }
  return(data)
}

#layout1<-encoding1(8)
#layout2<-encoding1(8)


### 2) 
crossover<-function(x,y){
  p<-sample(1:(length(x)/2),1)
  kid<-c(x[1:p],y[(p+1):length(x)])
  
  #while(any(duplicated(kid)==TRUE)){
  #  p<-sample((0:(length(x)/2)),1)
  #  kid<-c(x[1:p],y[(p+1):length(x)])
  #}
  return(kid)
}

#kid1<-crossover(layout1,layout2)

### 3)

mutate<-function(x){
  choice<-sample(length(x),1)
  a<-x[[choice]]
  x[[choice]]<-c(sample(length(x),1),sample(length(x),1))
  while(any(duplicated(x))== TRUE & all(x[[choice]] == a)){ #To ensure that the queen moves to a positon that is not filled by another queen
    x[[which(duplicated(x))]]<-c(sample(length(x),1),sample(length(x),1))
  }
  return(x)
}

#mutate(layout1)

### 4) 

fitness_binary<-function(x){
  a<-1
  
  #row check
  row<-c()
  for(i in 1:length(x)){
    row[i]<-x[[i]][1]
  }
  
  #column check
  col<-c()
  for(i in 1:length(x)){
    col[i]<-x[[i]][2]
  }
  
  if(any(duplicated(row))==TRUE){
    a<-0 #O implies that it is not a solution 
  }
  
  if(any(duplicated(col))==TRUE){
    a<-0
  }
  
  #diagonal check
  for(i in 1:length(x)){
    for(j in 1:length(x)){
      if(row[i]==row[j] & col[i]==col[j]){
        next
      }
      if(abs(row[i]-row[j])==abs(col[i]-col[j])){
        a<-0
      }
    }
  }

  return(a)
}

#fitness_binary(layout1)

fitness_attack<-function(x){
  dim<-combn(length(x),2)
  
  row<-c()
  for(i in 1:length(x)){
    row[i]<-x[[i]][1]
  }

  col<-c()
  for(i in 1:length(x)){
    col[i]<-x[[i]][2]
  }
  
  #non_attack<-length(x)
  count<-c()
  for(i in 1:ncol(dim)){
    if(row[dim[1,i]]==row[dim[2,i]] | col[dim[1,i]]==col[dim[2,i]] | abs(row[dim[1,i]]-row[dim[2,i]])==abs(col[dim[1,i]]-col[dim[2,i]])){
        count[i]<-1
    }
  }
  attack<-dim[,which(count==1)]
  a<-ncol(dim)
  if(any(dim(attack)>0)==TRUE){
    a<-abs(length(unique(c(unique(attack[1,]),unique(attack[2,]))))-length(x))
  }
  return(a)
}


#fitness_attack(layout1)

fitness_pair<-function(x){
  dim<-combn(length(x),2)
  
  row<-c()
  for(i in 1:length(x)){
    row[i]<-x[[i]][1]
  }
  
  col<-c()
  for(i in 1:length(x)){
    col[i]<-x[[i]][2]
  }
  
  #non_attack<-length(x)
  count<-c()
  for(i in 1:ncol(dim)){
    if(row[dim[1,i]]==row[dim[2,i]] | col[dim[1,i]]==col[dim[2,i]] | abs(row[dim[1,i]]-row[dim[2,i]])==abs(col[dim[1,i]]-col[dim[2,i]])){
      count[i]<-1
    }
  }
  number<-sum(count,na.rm = TRUE)
  a<-factorial(length(x))/(factorial(length(x)-2)*2)
  return(a-number)
}

#fitness_pair(layout1)


#For plotting
queens_attack<-function(x){
  dim<-combn(length(x),2)
  
  row<-c()
  for(i in 1:length(x)){
    row[i]<-x[[i]][1]
  }
  
  col<-c()
  for(i in 1:length(x)){
    col[i]<-x[[i]][2]
  }
  count<-c()
  for(i in 1:ncol(dim)){
    if(row[dim[1,i]]==row[dim[2,i]] | col[dim[1,i]]==col[dim[2,i]] | abs(row[dim[1,i]]-row[dim[2,i]])==abs(col[dim[1,i]]-col[dim[2,i]])){
      count[i]<-1
    }
  }
  number<-sum(count,na.rm = TRUE)
  return(number)
}

#queens_attack(layout1)


### 5), 6), 7)

genetic_algo<-function(encoding,mutation,fitness,n){
  population<-list()
  fitness_value<-c()
  for(i in 1:100){
    population[[i]]<-encoding(n)
    fitness_value[i]<-fitness(population[[i]])
  }
  
  plot_values<-c()
  for(i in 1:200){
    #print(i)
    worst_index <- which.min(fitness_value)
    
    #best<-population[[tail(order(fitness_value),1)]]
    best<-population[[which.max(fitness_value)]]
    #plot_values[i]<-queens_attack(best)
    
    
    parent<-sample(1:100,2)
    victim<-population[[order(fitness_value)[1]]]
    parent1<-population[[parent[1]]]
    parent2<-population[[parent[2]]]
    
    kid<-crossover(parent1,parent2)
    
    if(runif(1,0,1)>=mutation){
      kid<-mutate(kid)
    }
    fitness_kid <- fitness(kid)
    
    population[[worst_index]] <- kid
    fitness_value[worst_index] <- fitness_kid
    
    plot_values[i]<-queens_attack(kid)
    
    if(queens_attack(kid)==0){
      plot(plot_values,type="l")
      return("legal configuration found")
    }
    #population[[order(fitness_value)[1]]]<-kid 
    #fitness_value[order(fitness_value)[1]]<-fitness_pair(kid)
  }
  plot(plot_values,type="l")
  #return(queens_attack(kid))
}

```


### 8) 
```{r}
genetic_algo(encoding1,0.8,fitness_binary,4) 
genetic_algo(encoding1,0.8,fitness_binary,4)
genetic_algo(encoding1,0.5,fitness_binary,8)
genetic_algo(encoding1,0.5,fitness_binary,16)
genetic_algo(encoding1,0.1,fitness_binary,8)
genetic_algo(encoding1,0.9,fitness_binary,8)


genetic_algo(encoding1,0.5,fitness_attack,4) 
genetic_algo(encoding1,0.5,fitness_attack,8)
genetic_algo(encoding1,0.5,fitness_attack,16)
genetic_algo(encoding1,0.1,fitness_attack,8)
genetic_algo(encoding1,0.9,fitness_attack,8)


genetic_algo(encoding1,0.5,fitness_pair,4) 
genetic_algo(encoding1,0.5,fitness_pair,8)
genetic_algo(encoding1,0.5,fitness_pair,16)
genetic_algo(encoding1,0.1,fitness_pair,8)
genetic_algo(encoding1,0.9,fitness_pair,8)
```
Couldn't find legal configuration for all via the first encoding when the number of iterations is 100. But after increasing the interations to 200, there are some that give the legal configuration. As we increase the value of "n", it becomes harder to find the legal configuration or we do get it if we keep running the function multiple times until a legal configuration is found. So we need to run the same function with the same inputs multiple times to 
get the legel configuration. 



Encoding-2

```{r echo=TRUE}
toBits <- function (x, nBits = 8){
  tail(rev(as.numeric(intToBits(x))),nBits)
}

#toBits(8,5)

### 1)

### b)

encoding2<-function(n){
  index<-sample(n,n,replace = FALSE)
  output<-list()
  for(i in 1:n){
    #output[[i]]<-rep(0,n)
    output[[i]]<-toBits(index[i],log2(n))
  }
  return(output)
}

#e1<-encoding2(8)
#e2<-encoding2(8)

#To convert bits to integer 
bits_convert<-function(x){
  b<-c()
  for(i in 1:length(x)){
    c<-""
    for(j in 1:length(x[[1]])){
      c<-paste(c,x[[i]][j],sep = "")
    }
    b[i]<-unbinary(c)
  }
  b[b==0]<-length(x)
  return(b)
}

#bits_convert(e1)

#unbinary("111")
#b<-c()
#for(i in 1:8){
#  b[i]<-unbinary(paste(e1[[i]][1],e1[[i]][2],e1[[i]][3],sep = ""))
#}
# 0 implies 8 


### 2)
crossover2<-function(x,y){
  kid<-list()
  for(i in 1:length(x)){
    p<-sample(1:round(log2(length(x))/2),1)
    kid[[i]]<-c(x[[i]][1:p],y[[i]][(p+1):log2(length(x))])
  }
  return(kid)
}

#crossover2(e1,e2)

### 3)
mutate2<-function(x){
  choice<-sample(length(x),1)
  value<-sample(length(x),1)
  x[[choice]]<-toBits(value,log2(length(x)))
  
  return(x)
}

#mutate2(e1)


### 4)
fitness_binary_2<-function(x){
  x<-bits_convert(x)
  
  dim<-combn(length(x),2)
  
  a<-1 # if it is a legal config 
  if(any(duplicated(x))==TRUE){
    a<-0 
  }
  
  count<-c()
  for(i in 1:ncol(dim)){
    if(abs(x[dim[1,i]]-x[dim[2,i]])==abs(dim[1,i]-dim[2,i])){
      count[i]<-0
    }else{
      count[i]<-1
    }
  }
  if(any(count==0)){
    a<-0
  }
  
  return(a)
}

fitness_attack_2<-function(x){
  x<-bits_convert(x)
  dim<-combn(length(x),2)
  count<-c()
  for(i in 1:ncol(dim)){
    if(abs(x[dim[1,i]]-x[dim[2,i]])==abs(dim[1,i]-dim[2,i])){
      count[i]<-1
    }
  }
  attack<-dim[,which(count==1)]
  a<-ncol(dim)
  if(any(dim(attack)>0)==TRUE){
    a<-abs(length(unique(c(unique(attack[1,]),unique(attack[2,]))))-length(x))
  }
  return(a)
}

fitness_pair_2<-function(x){
  x<-bits_convert(x)
  dim<-combn(length(x),2)
  count<-c()
  for(i in 1:ncol(dim)){
    if(abs(x[dim[1,i]]-x[dim[2,i]])==abs(dim[1,i]-dim[2,i])){
      count[i]<-1
    }
  }
  
  number<-sum(count,na.rm = TRUE)
  a<-factorial(length(x))/(factorial(length(x)-2)*2)
  return(a-number)
}


#For plotting
queens_attack_2<-function(x){
  x<-bits_convert(x)
  dim<-combn(length(x),2)
  
  count<-c()
  for(i in 1:ncol(dim)){
    if(abs(x[dim[1,i]]-x[dim[2,i]])==abs(dim[1,i]-dim[2,i])){
      count[i]<-1
    }
  }
  
  number<-sum(count,na.rm = TRUE)
  return(number)
}

### 5), 6), 7)
genetic_algo_2<-function(encoding,mutation,fitness,n){
  population<-list()
  fitness_value<-c()
  for(i in 1:100){
    population[[i]]<-encoding(n)
    fitness_value[i]<-fitness(population[[i]])
  }
  
  plot_values<-c()
  for(i in 1:200){
    #print(i)
    worst_index <- which.min(fitness_value)
    
    #best<-population[[tail(order(fitness_value),1)]]
    best<-population[[which.max(fitness_value)]]
    #plot_values[i]<-queens_attack(best)
    
    
    parent<-sample(1:100,2)
    victim<-population[[order(fitness_value)[1]]]
    parent1<-population[[parent[1]]]
    parent2<-population[[parent[2]]]
    
    kid<-crossover2(parent1,parent2)
    
    if(runif(1,0,1)>=mutation){
      kid<-mutate2(kid)
    }
    fitness_kid <- fitness(kid)
    
    population[[worst_index]] <- kid
    fitness_value[worst_index] <- fitness_kid
    
    plot_values[i]<-queens_attack_2(kid)
    
    if(queens_attack_2(kid)==0){
      plot(plot_values,type="l")
      return("legal configuration found")
    }
    #population[[order(fitness_value)[1]]]<-kid 
    #fitness_value[order(fitness_value)[1]]<-fitness_pair(kid)
  }
  plot(plot_values,type="l")
  #return(queens_attack(kid))
}

```


8)
```{r}
genetic_algo_2(encoding2,0.8,fitness_binary_2,4)
genetic_algo_2(encoding2,0.5,fitness_binary_2,8)
genetic_algo_2(encoding2,0.5,fitness_binary_2,16)
genetic_algo_2(encoding2,0.1,fitness_binary_2,8)
genetic_algo_2(encoding2,0.9,fitness_binary_2,8)


genetic_algo_2(encoding2,0.5,fitness_attack_2,4) 
genetic_algo_2(encoding2,0.5,fitness_attack_2,8)
genetic_algo_2(encoding2,0.5,fitness_attack_2,16)
genetic_algo_2(encoding2,0.1,fitness_attack_2,4)
genetic_algo_2(encoding2,0.9,fitness_attack_2,4)


genetic_algo_2(encoding2,0.5,fitness_pair_2,4) 
genetic_algo_2(encoding2,0.5,fitness_pair_2,8)
genetic_algo_2(encoding2,0.5,fitness_pair_2,16)
genetic_algo_2(encoding2,0.1,fitness_pair_2,8)
genetic_algo_2(encoding2,0.9,fitness_pair_2,8)

```

We find more legal configurations for this for different input changes as compared to encoding-1. We can sometimes repeat the same function multiple times to get the legal configuration if not found in the first try. 

Encoding-3

```{r echo=TRUE}
### 1)

### c)

encoding3<-function(n){
  output<-sample(n,n,replace = FALSE)
  return(output)
}

l1<-encoding3(8)
l2<-encoding3(8)

### 2)

crossover3<-function(x,y){
  p<-sample(1:(length(x)/2),1)
  kid<-c(x[1:p],y[(p+1):length(x)])
  return(kid)
}

#crossover3(l1,l2)

### 3)
mutate3<-function(x){
  choice<-sample(length(x),1)
  x[choice]<-sample(length(x),1)
  return(x)
}

#mutate3(l1)

### 4)
fitness_binary_3<-function(x){
  dim<-combn(length(x),2)
  
  a<-1 # if it is a legal config 
  if(any(duplicated(x))==TRUE){
    a<-0 
  }
  
  count<-c()
  for(i in 1:ncol(dim)){
    if(abs(x[dim[1,i]]-x[dim[2,i]])==abs(dim[1,i]-dim[2,i])){
      count[i]<-0
    }else{
      count[i]<-1
    }
  }
  if(any(count==0)){
    a<-0
  }
  
  return(a)
}


fitness_attack_3<-function(x){
  dim<-combn(length(x),2)
  count<-c()
  for(i in 1:ncol(dim)){
    if(abs(x[dim[1,i]]-x[dim[2,i]])==abs(dim[1,i]-dim[2,i])){
      count[i]<-1
    }
  }
  attack<-dim[,which(count==1)]
  a<-ncol(dim)
  if(any(dim(attack)>0)==TRUE){
    a<-abs(length(unique(c(unique(attack[1,]),unique(attack[2,]))))-length(x))
  }
  return(a)
}


fitness_pair_3<-function(x){
  dim<-combn(length(x),2)
  count<-c()
  for(i in 1:ncol(dim)){
    if(abs(x[dim[1,i]]-x[dim[2,i]])==abs(dim[1,i]-dim[2,i])){
      count[i]<-1
    }
  }
  
  number<-sum(count,na.rm = TRUE)
  a<-factorial(length(x))/(factorial(length(x)-2)*2)
  return(a-number)
}

#fitness_pair_3(l1)


#For plotting
queens_attack_3<-function(x){
  dim<-combn(length(x),2)
  
  count<-c()
  for(i in 1:ncol(dim)){
    if(abs(x[dim[1,i]]-x[dim[2,i]])==abs(dim[1,i]-dim[2,i])){
      count[i]<-1
    }
  }
  
  number<-sum(count,na.rm = TRUE)
  return(number)
}

#queens_attack_3(l1)

### 5), 6), 7)
genetic_algo_3<-function(encoding,mutation,fitness,n){
  population<-list()
  fitness_value<-c()
  for(i in 1:100){
    population[[i]]<-encoding(n)
    fitness_value[i]<-fitness(population[[i]])
  }
  
  plot_values<-c()
  for(i in 1:200){
    #print(i)
    worst_index <- which.min(fitness_value)
    
    #best<-population[[tail(order(fitness_value),1)]]
    best<-population[[which.max(fitness_value)]]
    #plot_values[i]<-queens_attack(best)
    
    
    parent<-sample(1:100,2)
    victim<-population[[order(fitness_value)[1]]]
    parent1<-population[[parent[1]]]
    parent2<-population[[parent[2]]]
    
    kid<-crossover3(parent1,parent2)
    
    if(runif(1,0,1)>=mutation){
      kid<-mutate3(kid)
    }
    fitness_kid <- fitness(kid)
    
    population[[worst_index]] <- kid
    fitness_value[worst_index] <- fitness_kid
    
    plot_values[i]<-queens_attack_3(kid)
    
    if(queens_attack_3(kid)==0){
      plot(plot_values,type="l")
      return("legal configuration found")
    }
    #population[[order(fitness_value)[1]]]<-kid 
    #fitness_value[order(fitness_value)[1]]<-fitness_pair(kid)
  }
  plot(plot_values,type="l")
  #return(queens_attack(kid))
}
```

8)
```{r}
genetic_algo_3(encoding3,0.5,fitness_binary_3,4)
genetic_algo_3(encoding3,0.5,fitness_binary_3,8)
genetic_algo_3(encoding3,0.5,fitness_binary_3,16)
genetic_algo_3(encoding3,0.1,fitness_binary_3,8)
genetic_algo_3(encoding3,0.9,fitness_binary_3,8)


genetic_algo_3(encoding3,0.5,fitness_attack_3,4) 
genetic_algo_3(encoding3,0.5,fitness_attack_3,8)
genetic_algo_3(encoding3,0.5,fitness_attack_3,16)
genetic_algo_3(encoding3,0.1,fitness_attack_3,4)
genetic_algo_3(encoding3,0.9,fitness_attack_3,4)


genetic_algo_3(encoding3,0.5,fitness_pair_3,4) 
genetic_algo_3(encoding3,0.5,fitness_pair_3,8)
genetic_algo_3(encoding3,0.5,fitness_pair_3,16)
genetic_algo_3(encoding3,0.1,fitness_pair_3,8)
genetic_algo_3(encoding3,0.9,fitness_pair_3,8)

```

We find the most legal configurations for this for different input changes. We can sometimes repeat the same function multiple times to get the legal configuration if not found in the first try. 


9)
In terms of selecting which encoding would be the best, encoding 2 and 3 worked best to help find the legal configuration. But if we 
compare encoding 2 and 3, encoding 3 seems to provide more legal configurations based on the above inputs. 

In terms of the fitness/objective function, the binary function doesn't work well as it just provides an output of 1 and 0, and doesn't 
say much about how good or bad each chessboard layout is so it isn't a very useful fitness function in this case. Even though we do get 
some legal configuration using this, the other fitness functions get the legal configuration in lesser iterations. While both the "number of queens not attacked" function and the "pairs of queens attacking each other" function seem to perform equally good to help get the legal configuration. 

