---
title: "Q1 Rmd"
author: "Simon Jorstedt"
date: "`r Sys.Date()`"
output: pdf_document
---


# Question 1
We are given a gamma distribution with density

$$f(x) \propto x^5 e^{-x}, \space \space x > 0$$

## Question 1.a, 1.b, 1.c
We will implement the Metropolis-Hastings algorithm as a function where the proposal distribution can be specified. We will then run the algorithm to generate $10^4$ samples each for the Log-Normal $LN(X_t, 1)$, the chi-squared $\chi^2_{\lfloor X_t+1 \rfloor}$, and the THIS distributions. The result of this is reported in Figures 1, 2 and 3 respectively.

```{r Metro_Hasting function, echo=F}
# Question 1.a, 1.b, 1.c

# Define a function proportional to the target density
f_propo <- function(x){if(all(x >= 0)){(1/120)*x^5 * exp(-x)} else{NA}}

# Create proposal density functions
lognorm_proposal <- function(x=NA, param, sample=F){
  if(sample == F){dlnorm(x, meanlog = param, sdlog = 1)}
  else{rlnorm(1, meanlog = param)}
}

chi_proposal <- function(x=NA, param, sample=F){
  if(sample == F){dchisq(x, df = floor(param+1))}
  else{rchisq(1, df=floor(param+1))}
}

new_proposal <- function(x=NA, param, sample=F){
  if(sample == F){dexp(x, rate=1/param)}
  else{rexp(1, rate=1/param)}
}


# Metro_Hasting Instructions
# 
# x_start is a 1-dim numeric.
# prop_dens is a function which takes one named argument x,
# and one unnamed argument, which is supposed to fit in with the distribution.
# n is numeric, indicating number of observations to take (including burn-in)
Metro_Hasting <- function(x_start, prop_dens, n=1) {
  # Setup
  x_t <- x_start
  x_values <- c()

  repeat{
    # Generate new proposal
    x_new <- prop_dens(param=x_t, sample = T)
    #x_new <- rlnorm(1, meanlog = x_t, sdlog = 1)
    
    # Calculate probability value a
    # as product of Bayesian proposal ratio
    # and proposal density ratio (g(x_t | x_new) / g(x_new | x_t))
    Bay_pos_rat <- (f_propo(x_new) / f_propo(x_t))
    prop_ratio <- prop_dens(x=x_t, x_new) / prop_dens(x=x_new, x_t)
    #prop_ratio <- dlnorm(x_t, meanlog = x_new) / dlnorm(x_new, meanlog = x_t)
    a =  Bay_pos_rat * prop_ratio
    
    # Evaluate proposal value
    # Accept the new value if u < a (even if a >= 1)
    # Reject the new value with probability 1-a (if a < 1)
    u <- runif(1)
    #print(u)
    #print(a)
    if(u < a){x_t <- x_new}
    
    # Store the new value (even if x_t has not changed)
    x_values <- c(x_values, x_t)
    
    # Check break condition
    if(length(x_values) >= 10^4){break}
  }
  
  return(x_values)
}
```

## Question 1.d

```{r Figure creation, echo=F, fig.cap="Fig test"}
# Question 1.d

# Generate Figure 1: Lognormal(x_t, 1) as proposal
## Generate x values
set.seed(43849664)
x_values_1 <- Metro_Hasting(3, prop_dens = lognorm_proposal, n = 10^4)

## Plot the chain, histogram and combine them
par(mfrow=c(1,2))
plot(x_values_1, type = "l")
hist(x_values_1[-c(1:51)], main="Burn-in values", freq = F)
points(seq(0, 20, length=100), f_propo(seq(0, 20, length=100)), type="l")



# Generate Figure 2: Lognormal(x_t, 1) as proposal
## Generate x values
set.seed(9427452)
x_values_2 <- Metro_Hasting(3, prop_dens = chi_proposal, n = 10^4)

## Plot the chain, histogram and combine them
par(mfrow=c(1,2))
plot(x_values_2, type = "l")
hist(x_values_2[-c(1:51)], main="Burn-in values", freq=F)
points(seq(0, 20, length=100), f_propo(seq(0, 20, length=100)), type="l")



# Generate Figure 3: Normal(x_t, sd=1) as proposal
## Generate x values
set.seed(971614)
x_values_3 <- Metro_Hasting(5, prop_dens = new_proposal, n = 10^4)

## Plot the chain, histogram and combine them
par(mfrow=c(1,2))
plot(x_values_3, type = "l")
hist(x_values_3[-c(1:51)], main="Burn-in values", freq=F)
points(seq(0, 20, length=100), f_propo(seq(0, 20, length=100)), type="l")
```


## Question 1.e, 1.f
One can estimate the expected value from each of the samples by averaging over them, for the log normal proposal, this yields `r round(mean(x_values_1[-c(1:51)]), 2)`. For the chi-squared proposal, this yields `r round(mean(x_values_2[-c(1:51)]), 2)`. For the exponential proposal, this yields `r round(mean(x_values_3[-c(1:51)]), 2)`. The Gamma distribution of interest can quickly be identified as having parameters $k=6$, and $\theta = 1$, and thus expected value $6$.






# Old code
```{r, echo=F, eval=F}
# Not the density
f_prop <- function(x){if(x > 0){x^5 * exp(-x)} else{NA}}

# Start
x_t <- 2.5
x_values <- c()

repeat{
  # Generate new proposal
  x_new <- rlnorm(1, meanlog = x_t, sdlog = 1)
  
  # Calculate probability value a
  # as product of Bayesian proposal ratio
  # and proposal density ratio (g(x_t | x_new) / g(x_new | x_t))
  Bay_pos_rat <- (f_prop(x_new) / f_prop(x_t))
  prop_ratio <- (dlnorm(x_t, meanlog = x_new) / dlnorm(x_new, meanlog = x_t))
  a =  Bay_pos_rat * prop_ratio
  
  # Evaluate proposal value
  # Accept the new value if u < a (even if a >= 1)
  # Reject the new value with probability 1-a (if a < 1)
  u <- runif(1)
  if(u < a){x_t <- x_new}
  
  # Store the new value (even if x_t has not changed)
  x_values <- c(x_values, x_t)
  
  # Check break condition
  if(length(x_values) >= 10^4){break}
}
```




