---
title: "Lab 3 report"
author: "Simon Jorstedt & Siddhesh Sreedar"
date: "`r Sys.Date()`"
output:
  pdf_document: default
---


# Question 1





# Question 2
We will create a function for sampling from the double exponential (Laplace) distribution with location parameter $0$, and scale parameter $1$, with the density being

$$g(x) = \frac{1}{2} \cdot \exp \Big( -|x| \Big)$$

Below is our implementation `rlaplace` which samples from the $\text{Laplace}(\mu, \lambda)$ distribution using the [\textit{inverse distribution function (IDF) method}](https://en.wikipedia.org/wiki/Inverse_transform_sampling). The inverse distribution function $G^-1(x)$ was obtained through our [reference](https://en.wikipedia.org/wiki/Laplace_distribution) for the Laplace distribution. As per the IDF method, `rlaplace` first generates a uniformly random number from $U(0,1)$ for each desired observation. Then the IDF is applied to each such observation, so that the resulting observations are $Laplace(0,1)$ distributed.

We then proceed to generate $10^4$ approximately independent observations from `rlaplace`, and plot the result in Figure 2.1, along with the true density curve of $Laplace(0,1)$. The results are quite good, as the distribution of the simulated values closely matches the true density.

```{r Laplace generator function}
# Generate laplace variates using Inverse Distribution Function (IDF)
rlaplace <- function(n=1, location=0, scale=1){
  # Generate n observations from U(0,1)
  u <- runif(n)
  
  # Calculate (and return) corresponding IDF values
  location - scale*sign(u - 0.5)*log(1-2*abs(u-0.5))
}
```

```{r Figure 2.1 Laplace plot, echo=F}
# Laplace density function for use in the histogram below 
dlaplace <- function(x, location, scale){
  (scale / 2) * exp(-scale*abs(x-location))
}

# Plot a histogram of 10^4 Laplace observations
set.seed(03737693)
hist(rlaplace(10000, 0, 1), breaks = 100, freq = F,
     main = "Fig 2.1. Density of simulated Laplace observations",
     xlab = "x")

# Add true Laplace density curve
points(x = seq(-5, 5, 0.1),
       y = dlaplace(seq(-5, 5, 0.1), 0, 1),
       type = "l")
```

We will now proceed to use the Rejection Sampling Method (RSM) (see [reference](https://en.wikipedia.org/wiki/Rejection_sampling)) to generate standard normal variates from $N(0,1)$. For the envelope we will use the $Laplace(0, 1)$ distribution, scaled appropriately by a constant $a$. In order to find the optimal $a$, we construct the following expression and then derive the optimal (maximal) $a$ that satisfies the inequality.

$$h(x) = \frac{g(x)}{f(x)} = \frac{1/2 \cdot \exp(-x)}{1/\sqrt{2\pi} \cdot \exp(-\frac{x^2}{2})} = \sqrt{\frac{\pi}{2}} \cdot \exp \Big(\frac{x^2}{2} - x \Big) \geq a$$
$$h'(x) = \sqrt{\frac{\pi}{2}} \cdot (x-1) \cdot \exp \Big(\frac{x^2}{2} - x \Big)$$
It is clear that $h(x)$ grows very large as $x$ approaches $\pm \infty$, and thus the lone extrema at $x = 1$ is a global minima. This implies that the optimal $a$ is

$$a = h(1) = \sqrt{\frac{\pi}{2}} \cdot \exp \Big (\frac{1^2}{2} - 1 \Big) \approx 0.76 \space \space \space \space \text{(rounded down)}$$

Our envelope is thus

$$e(x) = g(x)/a.$$

We can now proceed to simulate $2000$ random numbers from $N(0,1)$ using the rejection sampling method. Below we implement this in `rnorm_rejection`. We also measure and output the average rejection rate when the function is run.

```{r Standard Normal generator function}
# Our derived optimal a
a <- sqrt(pi/2) * exp(-1/2)

# Generate standard normal variates using rejection sampling
rnorm_rejection <- function(n=1){ #mean=0, sd=1
  output_vector <- c()
  
  # Keep track of total number of attempts
  n_tries <- 0
  
  # Run the algorithm once for every n desired values
  for (i in 1:n){
    # Repeat the sampling procedure until a value is accepted
    repeat{
      n_tries <- n_tries + 1
      u <- runif(1, 0, 1)
      Y <- rlaplace(1, 0, 1)
      ratio <- dnorm(Y) / (dlaplace(Y, 0, 1) / a)
      if(u <= ratio){output_vector <- c(output_vector, Y);break}
    }
  }
  cat("The average rejection rate:", (n_tries-n)/n_tries, "\n")
  return(output_vector)
}
```

In Figure 2.2 below, we plot $2000$ standard normal variates sampled using `rnorm_rejection`, along with the true standard normal density (instead of values sampled from `rnorm`). The simulated distribution clearly resembles the standard normal, but it is not very smooth, which is desired.

```{r Figure 2.2 Normal Plot, echo=F}
# Plot a histogram of simulated Normal variates
set.seed(9432719)
hist(rnorm_rejection(2000), breaks=100, freq = F,
     main = "Fig 2.2. Laplace sampled Normal variates",
     xlab = "x")

# Add true standard normal density curve
points(x = seq(-5, 5, 0.1), y = dnorm(seq(-5, 5, 0.1), 0, 1), type = "l")
```

